---
title: "STAT331 Final Project - Data Cleaning"
author: "Thien An Tran, Tejasree Kandibanda, Matthew Huang, Chloe Anbarcioglu"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

# Project Proposal

For our project, we want to explore the relationship between murder rate and happiness score in each country per year. We hypothesize that there will be a negative correlation between the two meaning that the higher the murder rates, the lower happiness scores. This could potentially indicate that safety and security are significant factors in overall happiness.

# Setup

```{r setup}
library("tidyverse")
murder <- read_csv("data/murder_total_deaths.csv")
happiness <- read_csv("data/hapiscore_whr.csv")
population <- read_csv("data/pop.csv")
```

We got these data sets from *gapminder.org*.

> **population**: contains total population counts for each country per year.

> **murder**: contains total number of estimated deaths from interpersonal violence for each country per year.

> **happiness**: contains happiness score (converted to 0 to 100 scale to be in terms of percentage) for each country per year.

# Data Cleaning

In order to find the relationship between murder rates and happiness score for each country per year, we need to merge the total murders data set and population data set. That way, we can get the murder rate per 100K people.

## Value Conversion

However, we first have to convert all the values into numbers (i.e. fix cases such as `1.1k` to be `1100`).

We apply the convert_value function to both the total murders and population data set as they both contain values that need to be converted.

```{r}
convert_value <- function(val) {
  val <- as.character(val)
  
  multiplier <- case_when(
    str_detect(val, "k") ~ 1e3,
    str_detect(val, "M") ~ 1e6,
    str_detect(val, "B") ~ 1e9,
    TRUE ~ 1
  )
  
  numeric_value <- as.numeric(str_remove_all(val, "[kMB]"))
  
  return(numeric_value * multiplier)
}
```

### Converting Murder Counts to Numeric

```{r}
murder_clean <- murder |>
  select(country, `2005`:`2019`) |> 
  pivot_longer(cols = `2005`:`2019`,
               names_to = "year",
               values_to = "murder_count") |> 
  mutate(across(murder_count, ~convert_value(.)))

murder_clean
```

### Converting Population to Numeric

```{r}
population_clean <- population |>
  select(country, `2005`:`2019`) |>
  pivot_longer(cols = `2005`:`2019`,
               names_to = "year",
               values_to = "population") |>
  mutate(across(population, ~convert_value(.)))

population_clean
```

## Getting the Murder Rate for each Country and Year

After cleaning the total murders and population data set, we can proceed to merging them to get a data set of the murder rate per 100k people for each country and year.

```{r}
murder_pop_merged <- murder_clean |>
  inner_join(population_clean, by = c("country", "year"))

murder_rate_clean <- murder_pop_merged |>
  mutate(murder_rate_per_100k = (murder_count / population) * 100000)

murder_rate_clean
```

## Transforming Happiness Data Set

We use pivot longer to transform the happiness score data set so that it matches with the murder per 100k people data set we created. We can then merge the murder rate per 100k data set with the happiness score data set to get our final data set.

```{r}
happiness_clean <- happiness |>
  select(country, `2005`:`2019`) |>
  pivot_longer(cols = `2005`:`2019`,
               names_to = "year",
               values_to = "happiness_score") |>
   drop_na(happiness_score)

happiness_merged <- murder_rate_clean |>
  inner_join(happiness_clean, by = c("country", "year"))

happiness_merged
```

## Check for NAs
```{r}
happiness_merged |>
  summarise(across(everything(), ~sum(is.na(.))))
```

# Final Dataset Description
```{r}
glimpse(happiness_merged)
```

The final data set contains 1,820 rows and 6 columns.

The columns contains `country`, `year`, `murder_count`, `population`, `murder_rate_per_100k`, and `happiness_score`.

It provides a comprehensive overview of the murder rates and happiness scores across various countries and years. Each entry in the data set corresponds to a unique combination of a country and a year (ranging from 2005 to 2019).

## Output to CSV file
```{r}
# write.csv(happiness_merged, file="murder_happiness.csv", row.names=FALSE)
```


